Structuring College Access: The Market Segment Model and College Board Geomarkets

### 2025-04-25 text/back_end_KS.qmd
Dallas rq2 results

We also analyze student list orders collected through public requests to assess how the racial/ethnic and socioeconomic composition of included versus excluded prospects varies when filtering on particular Geomarkets within the Dallas-Fort Worth metropolitan area. Specifically, we analyze three list orders placed by a research university targeting prospects across 87 different Geomarkets across the country, including all six Geomarkets in the Dallas-Fort Worth Metropolitan area. Each order also filtered for prospects in the 2019-2020 graduating class and based on PSAT scores across three ranges: 1070-1180, 1190-1260, and 1270-1520. Similar to Chicago analyses, we focus on prospects within the middle PSAT score range (1190-1260) from the Dallas-Fort Worth metropolitan area to simulate Geomarket contributions and assess patterns of inclusion versus exclusion.

>>>>

We also analyze student list orders collected through public requests to assess how the racial/ethnic and socioeconomic composition of included versus excluded prospects varies when filtering on particular Geomarkets within the Dallas-Fort Worth metropolitan area. Specifically, we analyze three list orders placed by a research university targeting prospects across 87 different Geomarkets nationwide, including all six Geomarkets in the Dallas-Fort Worth Metropolitan area. Each order also filtered for prospects in the 2019-2020 graduating class and based on PSAT scores across three ranges: 1070-1180, 1190-1260, and 1270-1520. Similar to Chicago analyses, we focus on prospects within the middle PSAT score range (1190-1260) from the Dallas-Fort Worth metropolitan area to simulate Geomarket contributions and assess patterns of inclusion versus exclusion.

@fig-dallas-rq2-race presents the racial/ethnic composition of purchased student profiles within Geomarkets in the Dallas-Forthworth area. At the lowest PSAT score thresholds, notable differences emerge in the racial/ethnic composition of prospects within Geomarkets when compared to the overall demographic makeup of the metropolitan area in 2020 (see @fig-dallas-rq1-race). For example, the largest number of prospects in the 1070-1180 PSAT range were from the Collin & Rockwall Counties (57% White, 7% Black, 17% Hispanic, 14% Asian) and West of Dallas/Forth Worth Metroplex Geomarkets (61% White, 7% Black, 19% Hispanic, 7% Asian), both of which yield predominantly White prospect pools. On the other hand, the Dallas County (36% White, 17% Asian, 15% Black, 28% Hispanic) and the Irving, Arlington, and Grand Prairie (34% White, 13% Asian, 15% Black, 34% Hispanic) Geomarket pools have the greatest proportions of non-White prospects (although yielding fewer overall students). Interestingly, while the City of Dallas is the most racially/ethnically diverse Geomarket in the larger metropolitan area, its prospect pool in @fig-dallas-rq2-race includes larger proportion of Hispanic (41%) prospects but lower proportions of Asian (4%) and Black (9%) students comparatively.

These racial/ethnic composition patterns shift as PSAT scores increase. Shares of White prospects within Geomarkets remain relatively proportional to the overall 2020 metropolitan distribution in @fig-dallas-rq1-race. Howeover, prospect pools within Geomarkets from the middle range PSAT score (1190-1260) to the high PSAT score range (1270-1520) become disproportionately more Asian, while the representation of Hispanic and Black prospects decline substantially.

@fig-dallas-rq2-midSAT-race presents the race/ethnicity of all prospects targeted in the middle PSAT score range across Geomarket contributions. @fig-dallas-rq2-midSAT-race shows nearly all Geomarkets contributed to the pool of White prospects at rates nearly proportional to their overall contributions to all included prospects, with the exception of the West of Dallas/Fort Worth Metroplex Geomarket contributing a disproportionately larger share (29% versus 23% for all prospects). Larger disparities across Geomarket contributions were evident for Asian, Black, and Hispanic students. The Dallas County (20% versus 13% for all prospects) and Colin and Rockwall Counties (47% versus 37% for all prospects) Geomarkets contributed a disproportionate larger share of Asian students, whereas nearly all other Geomarkets contributed disproportionately smaller shares.

Representational patterns across Geomarkets are reversed for Black and Hispanic prospects in @fig-dallas-rq2-midSAT-race. All Geomarkets besides Collin and Rockwall Counties and West of Dallas/Fort Worth Metroplex contributed equal or larger proportions of Black and Hispanic prospects relative to their contributions to the overall prospect pool in the Dallas-Fort Woth metropolitan area. However, the City of Dallas Geomarket contributed the largest disproportionate share of Hispanic prospects (19% versus 11% for all prospects), whereas the Dallas Counties Geomarket contributed the largest disproportionate share of Black prospects (21% versus 13% for all prospects).

Contributions of each Geomarket to the pool of prospects based on first-generation college student status from the Dallas-Fort Worth metropolitan area are presented in @fig-dallas-rq2-midSAT-ses. Only 303 prospects of the 3,928 total in this middle PSAT score range pool were first-generation college students whose parents did not attend college. Again, all Geomarkets besides Collin and Rockwall Counties (14% versus 36% for all prospects) and West of Dallas/Fort Worth Metroplex (15% versus 24% for all prospects) contributed equal or larger proportions of first-generation college students (no college) relative to their contributions to the overall prospect pool. Disproportioantely larger contributions range from six percentage points by the Dallas Counties Geomarket (19% versus 13% for all prospects) to 11 percentage points by the City of Dallas Geomarket (22% versus 11% for all prospects). Contributions become more proportional for the pool of first-generation college students whose parents attended but did not complete their degree. Proportional balances across Geomarkets are also evident for the 3,069 prospects that are not first-generation college students, with the exception of Collin and Rockwall Counties Geomarket contributing the largest disproportinate share of not first-generation college students (40%) relative to the contributions to the overall prospect pool (36%).

Lastly, @fig-dallas-rq2-midSAT-combo presents Geomarket contributions to prospect pools by both first-generation college student status and race/ethnicity in the Dallas-Fort Worth metropolitan area. Among all White prospects (n=2,094) whose contact information was purchased in this middle PSAT score range, approximately 3% were first-generation college students whose parents did not attend college, 11% were first-generation college students whose parents attended some college but did not complete their degree, and 86% were not first-generation college students. @fig-dallas-rq2-midSAT-combo shows nearly all Geomarkets contribute a nearly proportionate share of prospects by first-generation college students relative to the overall pool. The only exception is the Irving, Arlington, & Grand Prairie Geomarket contributing a relatively larger share of White first-generation college students whose parents attended some college but did not complete their degree (21% versus 11% for all prospects).

Similar patterns of one or two Geomarkets disproportionately contributing larger shares of first-generation college students was evident across other racial/ethnic categories in @fig-dallas-rq2-midSAT-combo. For instance, City of Dallas (18%) and City of Fort Worth (24%) contributed larger proportions of first-generation college students whose parents did not attend college relative to the overall pool of all prospects (8%). On the other hand, the Irving, Arlington, & Grand Prairie (23%) and Dallas Counties (26%) Geomarkets contributed disproportionately larger proportions of Black students whose parents attended some college but did not complete their degree relative to the overall pool of all prospects (17%). For Asian prospects, City of Fort Worth and Irving, Arlington, & Grand Prairie Geomarkets contributed disproportionately larger shares of first-generation college students across both parents with no college and some college. City of Dallas (38%), City of Fort Worth (39%), Irving, Arlington, & Grand Prairie (32%), Dallas Counties (28%) all contributed larger shares of Hispanic students whose parents did not attend college relative to the overall prospect pool of these first-generation college students (23%), whereas only Dallas Counties contributed to larger shares of Hispanic students who parents attended some college but did not attain their degree (26% versus 20% for all prospects).
